---
title: "Notebook 01: Download Flow Data"
output: html_notebook
---

# Goal
In this Project, I am going to performed exploratory data analysis on HIV flow cytometry dataset. In this notebook, I used the code to get the information of dataset (ID: FR-FCM-ZZZK) from flow repository. In addition, the code will download the package I wrote, which includes important functions that would be used in the project.

# Setup environment

import needed library
```{r}
# FlowRepositoryR library is needed to download data from flow repository
library(FlowRepositoryR)
```

setup directory
```{r}
# The directory to store the dataset
datadir <- "/data/Flow_HIV"
```

In this project I have created additional helper functions and organized them into a package
```{r}
# install the helper package
library(devtools)
install_github("clintko/bios822FinalProjPackages", force = TRUE)

# import the library
library(bios822FinalProjPackages)

# Check if the package is installed.
hello()
```

# Summary information of Flow Repository

Below is the summary of the dataset:

- Repository:
    - Experiment: IDCRP's HIV Natural History Study 
    - ID: FR-FCM-ZZZK 
    - Primary Researcher: Nima Aghaeepour
- Purpose:
    - PFC analysis of 466 subjects enrolled in Infectious Disease Clinical Research Program's Natural History Study
- Link: http://flowrepository.org/id/FR-FCM-ZZZK
- Citation:
    - Aghaeepour N (2012) IDCRP's HIV Natural History Study data set.
    - Aghaeepour N (2012) Early immunologic correlates of HIV protection can be identified from computational analysis of complex multivariate T-cell flow cytometry assays

```{r}
# Search if the flow repository ID "FR-FCM-ZZZK" exist
flowRep.search("FR-FCM-ZZZK")
```

information of flow repository dataset FR-FCM-ZZZK
```{r}
# data summary of flow repository FR-FCM-ZZZK
dat_sum <- flowRep.get("FR-FCM-ZZZK"); cat("\n")
summary(dat_sum)
```

# Download the files

The command below downloads the dataset from flow repository. The size of files is around 60 Gb in total. For more information, the files can be downloaded from the website:   
http://flowrepository.org/experiments/20/download_ziped_files

```{r}
dat_sum  <- flowRep.get("FR-FCM-ZZZK")
dat_file <- download(dat_sum, dirpath = datadir)
```

# Read in the fcs files
Here I check the if the files are downloaded correctly from the flow repository. I have also stored the total dataset into a workspace. 

Number of fcs files; check if there are 466 fcs files
```{r}
# bash command to see how many fcs files in the folder
system("ls /data/Flow_HIV/*.fcs | wc -l")
```

print out the files in directory to double check if the files are downloaded currectly
```{r}
fclist <- dir(datadir)
fclist <- fclist[grepl("fcs", fclist)]
fclist
```

read in a whole set of fcs files 
```{r}
fs <- read.flowSet(file.path(datadir, fclist), transformation = FALSE)
```

Store the result
```{r}
# save the whole dataset into a workspace
save(fs, file=file.path(datadir, "FlowHIV_Raw.RData"))
```

In addition, I have read a subset of fcs files for demontration.
```{r}
# read in the first 10 files
fs_sub <- read.flowSet(file.path(datadir, fclist[1:10]), transformation = FALSE)
```

Store the result
```{r}
# save the sub-dataset into a workspace
save(fs_sub, file=file.path(datadir, "FlowHIV_Raw_subset10.RData"))
```
